\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-17 docs/parco13/abstract.tex
Added patch based numerics section

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it has been realized early
that it helps to aggregate degrees of freedom (DOF) at the element
level, as it has been done with high-order spectral element
\cite{TufoFischer99}, finite volume
\cite{paramesh, chombo, samrai}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.
\comment{Are you looking for references to parallel AMR codes?  Or more general AMR codes based on the FVM?}
\todo{No, I was looking for unstructured FV references.  Would you have any,
and can we add the block-AMR references later?}

@@

To enable hardware acceleration for parallel dynamic AMR, we like to
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989, be-le:1991,
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping, which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The Clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

In this paper we describe our design for the coupling of forest-of-octree AMR
with Clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and conclude with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% Clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf; \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I hav added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed.
\end{enumerate}

@@

Patch: contains the local coordinates and copies of neighboring ghost degrees
of freedom.  The layer of ghost copies is usually 2 points deep.

@@

@@

The authors acknowledge valuable discussion with Randy LeVeque, Marsha Berger,
and Hans-Petter Langtangen.  The leaf/patch paradigm was independently presented
by B.\ as part of a talk at the SCI Institute, Utah, in July 2011.

>>>>

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\clawpack}{{\sc Clawpack\xspace}}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, finite volume (element?)
\cite{???}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.  The finite volume method
has typically been implemented using a single degree of freedom per
cell on either structured \cite{ppm, clawpack, weno} or unstructured
meshes \cite{ddfv, ausm+, openfoam, }.  \comment{Are you looking for
  references to parallel AMR codes?  Or more general AMR codes based
  on the FVM?}  \todo{No, I was looking for unstructured FV
  references.  Would you have any, and can we add the block-AMR
  references later?} \comment{But most finite volume methods
  (structured or unstructured) only have a single degree of freedom
  per mesh cell (``element''?)  Only recently have people increased
  the order by extending the stencil (without increase the DOFs per
  mesh cell).}

@@

To enable hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989, be-le:1991,
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping (except in ghost cell regions),
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

In this paper we describe our design for the coupling of forest-of-octree AMR
with \clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and conclude with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% Clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC : Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf; \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I hav added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed.
\end{enumerate}

@@

For hyperbolic problems, we integrate the solution on a a single patch
using the wave propagation algorithm described by R. J. LeVeque and
implemented in {\sc Clawpack} \cite{le:2002, clawpack}.  We assume a
single degree of freedom per mesh cell and reconstruct a piecewise
constant solution to obtain left and right states at a mesh cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface. For
non-linear problems and systems, an approximate Riemann solver, such
as a Roe solver, is typically used. Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When subcycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.

Grid refinement and de-refinement also requires interpolation from coarse grids
to newly created fine grids, and the averaging of fine grid data to a newly created
underlying coarse grid.

@@

* sphere results\\
* swirl example

To come...

@@

The authors acknowledge valuable discussion with Randy LeVeque, Marsha Berger,
and Hans-Petter Langtangen.  We also acknowledge David Ketchison and the KAUST sponsored HPC$^3$
numerics workshop at which the initial phases of this project were first discussed.
The leaf/patch paradigm was independently presented
by B.\ as part of a talk at the SCI Institute, Utah, in July 2011.
